Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-04-04 Manuscript/manuscript.tex
Lot of copy and paste

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot. Figure~\ref{fig:DataBoxPlot}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio
\begin{figure}[h]
  \centering 
  \includegraphics{DataBoxPlot}
  \label{fig:DataBoxPlot}
\end{figure}

@@

@@

\subsection{GLMM}
%Include latex math symbols for the equations
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
%Does this part need to be changed?
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

@@

%Table should go to the results section
All Men's semi finals and finals data
\begin{center}
  \begin{tabular}{|c  c  c  c  c  c  c |}
   \hline 
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}

@@

\subsection{Rank-based comparison for Clustered}
%This should go to data section
This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

This dataset is quite small to perform any formal statistical analysis on, but
the data was expanded to include data for athletes who competed at national
competitions (many of which were held from May-July of 2022).  Once the data
was collected, each athlete was designated to be a cluster and the stage of
competition (national or international) was specified to be the group.  Next,
a Clustered Wilcoxon rank sum test was performed using the clusrank R package
\citep{Clusrank}.  Using a clustered rank sum test was chosen over a typical t 
test so as to avoid any assumptions about the distribution of population means.

%Citations needed
%Include math formulas and the test statistic
%Equation of test statistic is from papers cited by clusrank. Asymptootic dist
%Was found by __ referecne and the implementation is from the clusrank R package
Using the "clusWilcox.test" function, a grouped and clustered Wilcoxon rank sum
test was performed with the cluster set to be the Athlete and group set to Year.
The method was set to be "ds" which is used for the Data and Satten method, which
is preferable when the cluster sizes are informative \citep{Clusrank}.  We do not care that some
athletes made it further in either competition than another or that it should
not matter how elite an athlete is at the 110 meter hurdles once we get to the
national international competition level.  Cluster sizes vary from 2 observations
to 6.

@@

%This should go in the method section, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1
To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds. At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

Prior to the 2022 World Championship, there was a 0.0425\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.  That means that we would expect a time below the threshold to occur
once every 2,352 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 588 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section).  Discussion should show that random effect
%and standard error increase and that 2022 is special.  Now talk about Seiko
%timing equipment

@@

The p values produced by the clustered rank sum tests were 0.000155 for the
same athletes competing in the same year, and 0.001464 for the same athletes in
different years.  To put it simply, there is not a reasonable explanation for 
this.  
It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. 6 countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athlete's improvement in times was
discussed earlier in the paper, but this data shows that on average there was
%Use a reference in this part
improvement for British, Polish, French, Brazilian, and Spanish athletes.

It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon 
Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and batch
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics and Seiko 
may want to do further testing of their own to determine what is a human's physically
fastest reaction time, and if they choose to do so, they should start with Devon
Allen.

Since 1985 Seiko Holding Corporations has served every World Athletics Championship
as the official timer \citep{Seiko}.  Since 1985, the technology and the ability
to accurately predict measurements: long jump distances, false starts, total time,
reaction time, etc. have all dramatically improved.  Seiko did not start tracking
reaction time as an official measurement until 1999 in the men's and women's 100 meter
hurdles.  Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the world
of running outside of the Olympics.

%Citation needed in this section
Seiko's technology for detecting reaction times relies on the pressure that athletes
exert on the plate when they push off.  Their systems measure the time differential
between when the "gun" goes off to start the race and when the pressure changes 
\citep{Seiko}  If an athlete has a reaction time under 0.10 seconds, it is deemed a 
false start as it is considered that no athlete can react so quickly \citep{Seiko-Timing}.  
Thus, a reaction time of for example 0.05, suggests that the athlete predicted the gun and it
was luck that caused their abnormally high time.  At the World Championship level,
World Athletics and Seiko want to remove that element of luck and thus impose the 0.1
second barrier.

In 2013, Seiko upgraded their timing equipment for sprinting events (100 meter hurdles
falls under this catergory) \citep{WorldAthletics_2013}.  Since 1999, the highest reaction 
times in the men's 100 meter hurdle were recorded in 2013.  That is not to say that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.

>>>>

Data was copied from the World Athletics website and pasted into an Excel
spreadsheet. The data covers the men's 110 meter hurdles and the women's 100
meter hurdles from 1999 to 2022.  The variable of prediction interest is reaction
time, and the predicting variables are year, stage of competition (heats, 
semifinal, final), total time, and gender.  The data can be found in the appendix:.  
Factors such as name, country, and identification number were
not considered important for the purposes of this research besides the times for
Devon Allen and the other United States athletes mentioned above.  The variable "Stage"
refers to whether the observation occurred during the "heats" (Preliminary round) which
is denoted by "H" in the data, "S" for "semifinals" (Of which there are either 2 or 3 
heats each year), and "F" for finals. The variable "Gender" has levels "M" and "F" 
which stand for "male" and "female" respectively.  The variable "Batch" takes a 
numerical value and every heat of every round from 1999 to 2022 was assigned it's
own number.  This was added because of suspected issues that reaction time may be
correlated with other runners reacting quickly (one runner reacting quickly may
result in other runners reacting quickly).  There is no statistical signficance to
how the batches were labeled, they were ordered chronologically by meet but
reverse chronologically by year (Batch 1 is a preliminary Heat of 2022 and Batch
111 corresponds to the 1999 finals). The data that looked at in the paper
is best visualized using a box plot. Figure~\ref{fig:DataBoxPlot}.  Included are
box plots of the reaction times of the men's semi finals and finals pooled
together and just the finals.
%Fix quotations
%Dont need the titles will add captions, use width =5, height = 3 in aspect ratio
\begin{figure}[h]
  \centering 
  \includegraphics[width=\textwidth,height=4in]{DataBoxPlot}
  \caption{Boxplot of semifinals and finals data}\label{fig:DataBoxPlot}
\end{figure}

@@

This section was motivated from a qualitative analysis that started by looking
at how United States athletes reacted at the 2022 United States Track and Field
Championships. From June 23-26 2022, USA held its Track
and Field Championship to decide who to send to the World Championships at Hayward 
Field, the exact same venue as used for the World Track and Field Championship a month 
later.  Thus, a baseline is able to be established for the four USA athletes who 
competed in at least one round of both events: Trey Cunningham, Daniel Roberts, 
Grant Holloway, and Devon Allen. Every athlete reacted faster in all of the World 
Track and Field Championship races compared to the USA Track and Field 
Championship races. Devon Allen raced three times at the USA Track and
Field Championship and three times at the World Track and Field  Championship. 
His reaction times were significantly faster in the World Track and
Field Championship compared to the USA Track and Field Championship as were his
teamates.  This comparison suggests that the timing system used between the two
meets may have been meaningfully different to the point where Devon Allen was
disqualified more because of faulty equipment rather than because he reacted
too quickly.  This table below highlights the difference in reaction times for
Devon Allen, Trey Cunningham, Grant Holloway, and Daniel Roberts. "USA" denotes
the USA Track and Field Championships, "World" denotes the World Track and Field
Championships, "N/A" was used for any athlete that did not compete and thus
did not register a reaction time.  All numbers listed in the table are the reaction
time for the respective athletes in seconds.

This dataset is quite small to perform any formal statistical analysis on, but
the data was expanded to include data for athletes who competed at national
competitions (many of which were held from May-July of 2022).  Once the data
was collected, each athlete was designated to be a cluster and the stage of
competition (national or international) was specified to be the group.

@@

\subsection{GLMM}
%Include latex math symbols for the equations
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

@@

@@

%T, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1
To try and approximate the distribution of the reaction times, R code was written
to try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds.  The code re-paramertizes the gamma distribution to be in terms of
$\mu$ and shape rather than by $\alpha$ and $\beta$.

At a size of n=12,000,000, the calculated probability of a reaction
time being below 0.1 seconds was as follows for the various different data sets:

\subsection{Rank-based comparison for Clustered}

%Citations needed
%Include math formulas and the test statistic
%Equation of test statistic is from papers cited by clusrank. Asymptootic dist
%Was found by __ referecne and the implementation is from the clusrank R package

The test statistic of the DS method of the Wilcoxon rank-sum statistic is:
$Z = \frac{S - E(S)}{\sqrt{\hat{\text{Var}}(S)}}$ where 
$S = \mathbb{E}[W^* | X, \delta]$,
$W^* = \frac{1}{N} + \sum_{i=1}^{N} \delta_{i}^{*} R_{i}^{*}$, $R_{i}^{*}$ is
the rank of $X_{i}^{*}$.
The asymptomatic distribution of the Wilcoxon rank-sum statistic was found by
\citet{datta2005rank} and the R package "clusrank" allows for its implementation.
\citep{jiang2017wilcoxon}.

Using the "clusWilcox.test" function, a grouped and clustered Wilcoxon rank sum
test was performed with the cluster set to be the Athlete and group set to Year.
The method was set to be "DS" which is used for the Data and Satten method, which
is preferable when the cluster sizes are informative \citep{jiang2017wilcoxon}.  We do not care that some
athletes made it further in either competition than another or that it should
not matter how elite an athlete is at the 110 meter hurdles once we get to the
national international competition level.  Cluster sizes vary from 2 observations
to 6.

@@

%Table should go to the results section
All Men's semi finals and finals data
\begin{center}
  \begin{tabular}{|c  c  c  c  c  c  c |}
   \hline 
   Model & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   linear one & -1667 & 835.5 & 0.1529 & 0.0014 & N/A & N/A \\
   \hline
   linear year & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\ 
   \hline
   gamma one & -1756 & 880.2 & 6.554 & 0.0598 & N/A & N/A \\
   \hline
   gamma year & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   gamma one log & -1756 & 880.2 & -1.880 & 0.0091 & N/A & N/A \\
   \hline
   gamma year log & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\ [0.5ex]
   \hline
  \end{tabular}
  \end{center}
%Section 4.1: Gamma model results show the fitting with and without 2022,
%talk about p values and thresholds from the simulations (do not include the
%results in the methods section). 
Using the gamma function with link set to log (whose summary statistics can be
found in the last row of the above table), the probability of the function being
less than 0.1 seconds can be calculated to see if Allen's disqualification was an anomaly.
Prior to the 2022 World Championship, there was a 0.0425\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.  That means that we would expect a time below the threshold to occur
once every 2,352 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (The World Championship is biennial), every 588 years there
should be a time as extreme as Allen's.  This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

@@

%Discussion should show that random effect
%and standard error increase and that 2022 is special.  Now talk about Seiko
%timing equipment

%Citation needed in this section
Since 1985 Seiko Holding Corporations has served every World Athletics Championship
as the official timer \citep{Seiko}.  Since 1985, the technology and the ability
to accurately predict measurements: long jump distances, false starts, total time,
reaction time, etc. have all dramatically improved.  Seiko did not start tracking
reaction time as an official measurement until 1999 in the men's and women's 100 meter
hurdles.  Seiko regularly updates their equipment so that they provide cutting edge
technology to the World Track and Field Championships, the highest stakes in the world
of running outside of the Olympics. Seiko's technology for detecting reaction 
times relies on the pressure that athletes
exert on the plate when they push off.  Their systems measure the time differential
between when the "gun" goes off to start the race and when the pressure changes 
\citep{Seiko}  If an athlete has a reaction time under 0.10 seconds, it is deemed a 
false start as it is considered that no athlete can react so quickly 
\citep{Seiko-Timing}.  Thus, a reaction time of for example 0.05, suggests that 
the athlete predicted the gun and it was luck that caused their abnormally high 
time.  At the World Championship level, World Athletics and Seiko want to remove 
that element of luck and thus impose the 0.1 second barrier. In 2013, Seiko 
upgraded their timing equipment for sprinting events (100 meter hurdles falls 
under this category) \citep{WorldAthletics_2013}.  Since 1999, the highest reaction 
times in the men's 110-meter hurdle were recorded in 2013.  That is not to say that the higher 
times were caused by Seiko's equipment, but the two may be related.  It is worth noting that prior
to 2022, Seiko again upgraded its technology but for its jump management system.

@@

For the comparison between athletes who competed in 2019 and 2022, the p value
was 0.001464.  As Seiko timed both events, there should not be a signficant
difference if the timing equipment in both instances was working correctly.
Additionally, the p value was 0.000155 for the same athletes competing in the 
same year  To put it simply, there is not a reasonable explanation for 
this. It is inconceivable that so many athletes would improve such a significant
amount in such a small amount of time. 6 countries are represented in this 
comparison, which drastically decreases the chances of this issue being caused by
one specific country.  The United States athlete's improvement in times was
discussed earlier in the paper, but this data shows that on average there was
%Use a reference in this part
improvement for British, Polish, French, Brazilian, and Spanish athletes.

%Wrap up and Conclusion
It is worth noting that in the semi-finals of the World Track and
Field Championships; Allen's reaction time was 0.101 which is only 0.001 above
the legal limit.  What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun.  It seems unlikely that
Allen would be able to correctly predict the gun with such precision two times in a row,
which is the exact reason that the 0.10 second reaction time disqualification barrier was
imposed in the first place.  But Allen showed that at the Championship he was able to
react extremely fast two times, which may suggest skill rather than luck. Devon 
Allen is a very good athlete, as evidenced by his football career at the University of Oregon
and making it onto the Philadelphia Eagles practice squad this past year \citep{Hurley}.
Thus the 0.099 reaction time may not have been a product of Devon Allen
predicting the start of the race but rather a combination of a quick reaction 
and a possibly faulty sensor.

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and batch
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics did not
do anything about \citet{komi2009iaaf}'s study, but perhaps they should now so
that the story of Devon Allen is not repeated.
